---
title: "NWPD_Shooting_Incident"
author: "BQ"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Project Description:
*In this project, we will analyze a shooting incident involving the NYPD. The data used is publicly available at "https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic".*


```{r}
# Set seed for reproducibility
set.seed(42)


# Define required packages
required_packages <- c("tidyverse", "ggplot2", "rstudioapi", "readxl")

# Check which of the required package is not installed in users' machine
need_install <- required_packages[!(required_packages) %in% installed.packages()]

# Install the required packages if any of them are not already installed
if (length(need_install) > 0 ){
  install.packages(need_install)
}


# Load packages
lapply(required_packages, require, character.only = TRUE)
```

```{r}
# Getting data
nypd_raw_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD") %>%
  distinct() %>%
  drop_na()

df <- nypd_raw_data
```



```{r}
# Display data structure
str(df)
```
#### Summary of the data frame
```{r}
summary(df)
```
```{r}
head(df)
```

## Data exploration, data cleaning and trasnformation
```{r}
# Getting count of daily incidents
daily_count <- df %>%
  group_by(OCCUR_DATE) %>%
  summarise(INCIDENT_COUNT = n(), .groups = "drop")

ggplot(daily_count, aes(x = INCIDENT_COUNT)) + 
  geom_histogram(binwidth = 1, fill = "grey", color = "black", alpha = 0.5) +
  labs(title = "Daily incidents distribution",
       x = "Number of incidents per day",
       y = "Frequency") +
  theme_classic()

```

```{r}
df <- df %>%
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    Year = year(OCCUR_DATE),
    Month = factor(month(OCCUR_DATE, label = TRUE, abbr = TRUE), levels = month.abb),
    DayOfWeek = factor(wday(OCCUR_DATE, label = TRUE, abbr = TRUE), levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")),
    TimeOfDay = case_when(
      hour(OCCUR_TIME) >= 6 & hour(OCCUR_TIME) < 12 ~ "Morning",
      hour(OCCUR_TIME) >= 12 & hour(OCCUR_TIME) < 18 ~ "Afternoon",
      hour(OCCUR_TIME) >= 18 & hour(OCCUR_TIME) < 24 ~ "Evening",
      TRUE ~ "Night"
    )
  )

```

```{r}
# Checking the new data structure
str(df)
```

```{r}
summary(df)
```

## Grouping and Summarization

#### Total Incident for each year
We group the data by 'Year' to count total number of of incidents for each year. This summarized data will be used later to create a bar chart comparing the yearly totals.

```{r}
# Group by Year and count total incidents
yearly_incidents <- df %>%
  group_by(Year) %>%
  summarise(Incident_Count = n(), .groups = "drop")

# View the results
print(yearly_incidents)

```

```{r }
# Add a percentage change column for annotations
yearly_incidents <- yearly_incidents %>%
  mutate(Percent_Change = c(NA, (Incident_Count[2] - Incident_Count[1]) / Incident_Count[1] * 100))

# Create the combined bar and line chart
ggplot(yearly_incidents, aes(x = Year)) +
  # Bar chart for incident counts
  geom_bar(aes(y = Incident_Count, fill = as.factor(Year)), stat = "identity", width = 0.7, show.legend = FALSE) +
  # Line plot for percentage change
  geom_line(aes(y = Incident_Count, group = 1), color = "orange", size = 1) +
  geom_point(aes(y = Incident_Count), color = "black", size = 3) +
  # Add percentage labels on the line
  geom_text(aes(y = Incident_Count, label = ifelse(is.na(Percent_Change), "", paste0(round(Percent_Change, 1), "%"))), 
            vjust = -1.5, size = 5, color = "black", na.rm = TRUE) +
  # labels and theme
  labs(
    title = "Yearly Shooting Incidents (2022 vs 2023)",
    x = "Year",
    y = "Incident Count",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("2022" = "blue", "2023" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```

#### Month to Month comparison
We will group the data by the Year and Month and count the incident for each combination, than will plot a monthly trends for 2022 and 2023 as a two separate lines for comparison.
```{r}
# Group by Year and Month, and count incidents
monthly_counts <- df %>%
  group_by(Year, Month) %>%
  summarise(Incident_Count = n(), .groups = "drop") %>%

# View results
print(monthly_counts)
```

```{r}
# Create the line chart
ggplot(monthly_counts, aes(x = Month, y = Incident_Count, colour = as.factor(Year), group = Year)) +
  geom_line(size = 1) + 
  geom_point(size = 2) +
  labs(
    title = "Monthly Shooting incidents: 2022 vs 2023",
    x = "Months",
    y = "Incident Count",
    color = "Years"
  ) +
  scale_color_manual(values = c("2022" = "blue", "2023" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```
*The Blue line (2022) consistently shows higher incident counts thatn the red line (2023) accoss most months*
*The peak in 2022 occurs in July, with more than 200 incidents*
*2023, while having fewer incidents overall, also shows a slight peak in the summer months (July-August)*
*2023 shows consistently lower counts compared to 2022, with 29.6% decline overall*
*Interestingly, the counts for November and December are very close in both years, indicating a leveling off of the decline towards the end of the year*





#### Incident by day of the week, yearly comperison
```{r}
# Group by DayOfWeek and count incidents
day_of_week_analysis <- df %>%
  group_by(Year, DayOfWeek, TimeOfDay) %>%
  summarise(Incident_Count = n(), .groups = "drop")

# View the data
print(day_of_week_analysis)
```
```{r}
# Create the bar chart
ggplot(day_of_week_analysis, aes(x = TimeOfDay, y = DayOfWeek, fill = Incident_Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  facet_wrap(~ Year) +
  labs(
    title = "Shooting Incidents by Day, Time and Year",
    x = "Time of Day",
    y = "Day of the Week",
    fill = "Incident Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 14, face = "bold")
  )
```
*Incidents are consistently high during the night across most days of the week*
*Saturdays and Sundays show particularly high incidents during the night in both years, with strong activity in the evening as well, which could be due to social gatherings*
*The heatmap for 2023 is generally lighter color than 2022, indicating fewer incidents accross all times and days, which aligns with previous analyses*




#### Count of Incident per Borough, Monthly and Yearly comparison
```{r}
# Group data by Month, Year and BORO
boro_monthly <- df %>%
  group_by(Year, Month, BORO) %>%
  summarise(Incident_Count = n(), .groups = "drop")

# View Monthly data
print(boro_monthly)

```
```{r}
# Create faceted line plots for monthly trends
ggplot(boro_monthly, aes(x = Month, y = Incident_Count, color = as.factor(Year), group = Year)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ BORO) +
  labs(
    title = "Monthly Shooting incidents by Borough (2022 vs 2023)",
    x = "Month",
    y = "Incident Count",
    color = "Year"
  ) +
  scale_color_manual(values = c("2022" = "blue", "2023" = "red")) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```
*Across all boroughs, incidents tent to peak in the summer months (June - August) and this seasonal pattern is consistent in both years*
*Incidents in 2023 (red line) are consistently lower than in 2022 (blue line) across most months and boroughs, aligning with the overall decline*
*BROOKLYN and BRONX have the highest number of incidents compared to other boroughs, with noticeable peaks in July for both years*
*STATEN ISLAND consistently has the lowest number of incidents with almost no seasonal variations. also the difference between 2022 and 2023 is minimal, indicating stability in this borough*
*The decline in 2023 is evident in all boroughs, but the Bronx and Brooklyn contribute the most to the overall decline*


## Machine Learning Model
#### Linear Regression Model
We will build a Linear Regression model to predict the number of shooting incidents (Incident_Count) for each borough (BORO) in a a specific month (Month) and year (Year).
```{r}
# Aggregate data by BORO, Month, Year
data_model <- df %>%
  group_by(BORO, Year, Month) %>%
  summarise(Incident_Count = n(), .groups = "drop") %>%
  mutate(BORO = as.factor(BORO))

print(head(data_model))
```

#### Train-Test Split
We will split the data into training(80%) and testing(20%) subsets to evaluate the model's performance
```{r}
# Import caret library for spliting the data
install.packages("caret")
```

```{r}
library(caret)
```

```{r}
#Split data into training and testing sets
set.seed(42) # for reproducibility
train_index <- createDataPartition(data_model$Incident_Count, p=0.8, list = FALSE)
train_data <- data_model[train_index,] # the 80% portion
test_data <- data_model[-train_index,] # the remaining 20% portion

# View the size of the train and test datasets
cat("Training data:", nrow(train_data), "\nTesting data:", nrow(test_data))
```

#### Fit a Linear Regression Model
We will train a simple Linear Regression model using lm() function with BORO, Month, and Year as predictor
```{r}
# Fit the linear regression model
lm_model <- lm(Incident_Count ~ BORO + Month + Year, data = train_data)

# View the model summary
summary(lm_model)
```

#### Model Prediction
We will use the trained modelt to predict Incident_Count on the test data
```{r}
# Predict on the test data
predictions <- predict(lm_model, newdata = test_data)

# Combine the predections with actual values
result <- data.frame(
  Actual = test_data$Incident_Count,
  Predicted = predictions
)

print(head(result))

```
*Actual values are a bit higher than Predicted values in the 6 rows of data, we will quantify how close these predictions are to the actual values using MAE and MSE*

```{r}
# Calculate MAE and MSE
mae <- mean(abs(result$Actual - result$Predicted))
mse <- mean((result$Actual - result$Predicted)^2)

# view the evaluation
cat("MAE: ", mae, "\n")
cat("MSE: ", mse, "\n")

```
*MAE value suggests the model performs reasonably well but has room for improvement*
*MSE value highlights that there may still be occasional large errors in prediction*


#### Result Visualization
We will create a scatter plot to visualize the relationship between actual and predicted values

```{r}
# Scatter plot of actual vs. predicted values
ggplot(result, aes(x = Actual, y = Predicted)) + 
  geom_point(color = "blue") + 
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Actual vs Predicted Shooting Incidents",
    x = "Actual Incident Count",
    y = "Predicted Incident Count"
  )+
  theme_minimal()
```














